---
title: "Week 2: dplyr and census"
description: "First impressions, and getting the workflow running."
date: 2026-09-04
---

## What We Did
First we had a quiz. It was a little scary but I am not scared anymore.


## 1. The R Environment

- **Console**: run code interactively (e.g. `2 + 2`), but nothing is saved.
- **R Script (.R)**: saves code so you can rerun/edit later.
- **Quarto (.qmd)**: combines text, headings, R code, tables, and figures into one reproducible document — code lives inside chunks:

```{r}
2 + 2
```

## 2. Reading R Code

Think of code in terms of **objects** (store results), **verbs/functions** (perform actions), **arguments** (instructions), and **data** (what's being analyzed).

```{r}
#| eval: false
result <- filter(counties, population > 50000)
```

| Component | Example | Meaning |
|---|---|---|
| Object | `result` | Stores the result |
| Assignment | `<-` | Assigns result to object |
| Function | `filter()` | Action performed |
| Data | `counties` | Data being processed |
| Argument | `population > 50000` | Condition applied |

`<-` is the **assignment operator**: `x <- 10` creates object `x` holding 10.

## 3. R Packages

```{r}
#| eval: false
install.packages("dplyr")   # install once
library(dplyr)               # load each session
library(tidycensus)
```

`install.packages()` = get it on your computer; `library()` = load it for use.

## 4. Data Frames and Tibbles

A **data frame** is data in rows (observations) and columns (variables); a **tibble** is a cleaner-printing tidyverse version.

| Name | Income | MOE | MOE_pct |
|---|---|---|---|
| Adams | 78,975 | 3,334 | 9 |
| Berks | 74,617 | 1,191 | 12 |

## 5. Common dplyr Verbs

| Function | Purpose |
|---|---|
| `filter()` | Pick rows by condition |
| `select()` | Pick columns |
| `mutate()` | Create/modify columns |
| `arrange()` | Reorder rows (`desc()` for descending) |
| `group_by()` | Group for per-group analysis |
| `summarize()` | Reduce to summary stats |

```{r}
#| eval: false
filter(counties, population > 50000)
select(counties, NAME, population)
counties <- mutate(counties, income2 = income * 2)
arrange(counties, desc(income))

counties |>
  group_by(state) |>
  summarize(min_income = min(income, na.rm = TRUE))
```

## 6. Pipes: "And Then"

`|>` (modern) and `%>%` (older tidyverse) chain operations together:

```{r}
#| eval: false
counties |>
  filter(moe_pct > 10) |>
  arrange(desc(moe_pct)) |>
  select(NAME, moe_pct)
```

Reads as: start with counties → filter → arrange → select. Equivalent to writing separate `step1`, `step2`, `step3` objects, but more compact.

## 7. R Is Case Sensitive

`counties` ≠ `Counties`; `moe_pct` ≠ `MOE_pct`. Always match exact spelling/capitalization.

## 8. Census Data

- **Decennial Census**: every 10 years, full population count.
- **ACS (American Community Survey)**: sample-based (since 2005), gives detailed demographic/economic/housing data plus a **margin of error (MOE)** since it's an estimate, not a full count.

## 9. ACS Data Products

- **1-Year estimates**: larger areas/populations only.
- **5-Year estimates**: smaller areas (tracts, block groups, counties, places); e.g. a 2023 5-year estimate covers the 5-year period ending 2023.

## 10. Estimates and Margins of Error

Every ACS value pairs an **estimate** with an **MOE** indicating uncertainty:

| County | Income | MOE |
|---|---|---|
| Adams | $78,975 | $3,334 |
| Bucks | $104,200 | $1,502 |

Always report/interpret estimates alongside their MOE.

## 11. Census Variable IDs

`B19013_001E` → **B**(base table) **19013**(table number) **001**(line number) **E**(estimate). The MOE version ends in **M** (`B19013_001M`).

## 12. Geography in Census Data

Hierarchy: **State → County → Census Tract → Block Group → Block**. Not all datasets cover every level. Tract boundaries can shift over time — use **NHGIS** or **LTDB/LTDR** for longitudinal comparisons.

## 13. Blocks and Block Groups

**Blocks** are the smallest unit (Decennial only, little ACS data). **Block groups** (collections of blocks) allow detailed analysis but often carry large MOEs due to small sample sizes.

## 14. GEOIDs

Unique geographic identifier, e.g. `42101000100` = `42`(PA) + `101`(county) + `000100`(tract). Used to join Census data to boundary files.

## 15. Getting Census Data with tidycensus

```{r}
#| eval: false
library(tidycensus)

data <- get_acs(
  geography = "county",
  variables = c("B19013_001", "B01003_001"),
  state = "PA",
  year = 2023,
  survey = "acs5"
)
```

| Argument | Purpose |
|---|---|
| `geography` | Geographic level |
| `variables` | Census variable codes |
| `state` | State filter |
| `year` | Data year |
| `survey` | ACS product (e.g. `acs5`) |

## 16. Long vs. Wide Data

**Long**: one row per observation-variable pair (good for multi-variable tidyverse workflows).

| GEOID | NAME | VARIABLE | ESTIMATE | MOE |
|---|---|---|---|---|
| 42001 | Adams | B19013_001 | 78,975 | 3,334 |

**Wide**: each variable is its own column (easier to scan per variable).

| GEOID | NAME | B19013_001 | MOE |
|---|---|---|---|
| 42001 | Adams | 78,975 | 3,334 |

## 17. Key Concepts to Remember

- Console = testing; R script = saved code; Quarto chunk = code inside `.qmd`.
- Objects store data; functions act on objects; arguments guide functions.
- `filter()` rows, `select()` columns, `mutate()` new/modified columns, `arrange()` order, `group_by()`/`summarize()` grouped stats.
- Pipes (`|>`/`%>%`) chain steps; R is case sensitive.
- ACS = estimates with MOE, not exact counts; GEOIDs identify geography; geography can change over time.
- `tidycensus` fetches Census/ACS data; results can be long or wide format.
