---
title: "Week 1: Getting Set Up"
description: "First impressions, and getting the workflow running."
date: 2026-08-28
---

## What We Did


# What Is a Website?

A website is a collection of files and folders that a web browser can read and display. The main files are usually **HTML** files.

* **HTML** stands for HyperText Markup Language.
* A browser—such as Chrome, Safari, Edge, or Firefox—reads HTML and displays it as a web page.
* A website normally has one or more HTML pages, plus supporting files such as images, CSS stylesheets, JavaScript, fonts, and data files.

For this course, you generally do **not** write large amounts of HTML by hand. You write Quarto documents, and Quarto creates the HTML for you.

# The Quarto Workflow

The basic Quarto workflow looks like this:

```text
You write .qmd files
        ↓
Quarto reads and renders them
        ↓
Quarto runs R code where needed
        ↓
Quarto creates .html files
        ↓
GitHub Pages publishes the HTML online
```

A `.qmd` file is a **Quarto Markdown** document. It is designed to hold narrative writing, headings, links, images, equations, and executable code in one file.

Quarto is useful because it can produce web pages and other outputs—such as reports, presentations, PDFs, and Word documents—from a single source document.

# What Does "Render" Mean?

To **render** means to build the finished output from your source file.

When you render a Quarto document, Quarto generally:

1. Reads the `.qmd` file from top to bottom.
2. Interprets the YAML settings at the top.
3. Reads the Markdown text, headings, links, and images.
4. Finds executable code cells/chunks.
5. Sends R code to R, runs it, and receives the results.
6. Combines the prose, code output, tables, figures, and formatting.
7. Produces an HTML page for the website.

For example, this code in a `.qmd` document:

```{r}
2 + 2
```

will be evaluated during rendering, and the resulting webpage can display:

```text
[1] 4
```

Depending on the document or project settings, you can show or hide the code, warnings, messages, and results.

# Render Versus Preview

These are related but different actions:

| Action                      | What it does                                          | Who can see it?                                  |
| :-------------------------- | :---------------------------------------------------- | :----------------------------------------------- |
| **Save**                    | Saves your `.qmd` source file on your computer        | Only you                                         |
| **Render**                  | Converts `.qmd` into `.html`                          | Initially only you                               |
| **Preview**                 | Opens the rendered page locally, often at `localhost` | Only you                                         |
| **Commit**                  | Creates a labeled Git snapshot on your computer       | You and anyone with access to the local computer |
| **Push**                    | Uploads commits to GitHub                             | People with repository access                    |
| **GitHub Pages deployment** | Publishes the rendered HTML as a public website       | Anyone with the URL                              |

A URL such as `http://localhost:6930/about.html` is a local preview.

**"Localhost" means your own computer**, so an instructor cannot open that address.

Your public website is instead a GitHub Pages URL such as:

`https://username.github.io/repository-name/`

# Anatomy of a `.qmd` File

A Quarto document has three common parts:

1. YAML metadata at the top
2. Markdown content
3. Executable code cells/chunks

## 1. YAML

YAML is the configuration block at the top of a `.qmd` file. It begins and ends with three dashes:

```yaml
---
title: "My Portfolio"
author: "Emmanuel Sampson"
date: today
format: html
---
```

YAML can specify:

* The title and author
* The date
* Output format, such as HTML
* Table of contents settings
* Whether code is shown or hidden
* Other document-level options

### Important YAML Rules

* Indentation matters.
* Use spaces, not tabs.
* A colon must have a space after it:

```yaml
title: "My Page"
```

* YAML begins at the first line of the document.
* The opening and closing `---` lines must both be present.

A common error is incorrect indentation.

For example, this is **wrong**:

```yaml
website:
navbar:
  title: "Portfolio"
```

Because `navbar` should be nested under `website`.

The correct version is:

```yaml
website:
  navbar:
    title: "Portfolio"
```

# 2. Markdown

Markdown is the simple writing syntax used after YAML. It creates headings, paragraphs, lists, links, images, emphasis, and more.

For example:

```markdown
# Main Page Title

## Major Section

### Smaller Section

This is an ordinary paragraph.

- First bullet
- Second bullet

1. First numbered item
2. Second numbered item
```

## Useful Markdown Patterns

| What you want   | Markdown you write            | What it does                |
| :-------------- | :---------------------------- | :-------------------------- |
| Main heading    | `# Title`                     | Creates the largest heading |
| Section heading | `## Section`                  | Creates a major section     |
| Subsection      | `### Subsection`              | Creates a subsection        |
| Bold text       | `**important**`               | Makes text bold             |
| Italic text     | `*emphasis*`                  | Makes text italic           |
| Link            | `[text](https://example.com)` | Creates a clickable link    |
| Bullet list     | `- item`                      | Creates a bulleted list     |
| Numbered list   | `1. item`                     | Creates a numbered list     |
| Inline code     | `` `mean(x)` ``               | Displays code in text       |

## Markdown Example

```markdown
## What We Learned Today

We learned how **Quarto** turns text and R code into a webpage.

For more information, visit the [Quarto website](https://quarto.org).
```

# 3. Code Chunks/C ells

A code chunk is a fenced block that tells Quarto to run code.

In a Quarto document using R, a basic code cell looks like this:

```{r}
2 + 2
```

You can give a cell a meaningful label:

```{r load-tidycensus}
library(tidycensus)
```

Good chunk names help you find and debug code.

Use names that describe the purpose, such as:

* `load-packages`
* `download-census-data`
* `clean-data`
* `make-map`
* `plot-income-distribution`

**Avoid spaces in chunk labels.** Use hyphens instead.

## Code Chunk Options

You can add options to control what viewers see:

```r
2 + 2
```

Common options include:

* `echo: true` or `echo=true`: Shows the code in the webpage.
* `echo: false` or `echo=false`: Runs the code but hides it.
* `warning: false`: Hides warning messages.
* `message: false`: Hides package startup messages.
* `include: false`: Runs the code but shows neither the code nor the output.

# Files and Folders in a Quarto Website

A Quarto website contains files for people to read and files used by software to build and manage the project.

## Pages Written for People

| File or folder           | Purpose                                              |
| :----------------------- | :--------------------------------------------------- |
| `index.qmd`              | The home page of your site                           |
| `about.qmd`              | Your About page                                      |
| `contact.qmd`            | A Get in Touch/contact page, if you create one       |
| `weekly-notes/index.qmd` | The main page for weekly notes                       |
| `weekly-notes/`          | Folder holding weekly-note pages or materials        |
| `assignments/index.qmd`  | The main page listing assignments                    |
| `assignments/`           | Folder for assignment pages, data, code, and figures |
| `images/` or `img/`      | Images used in the website                           |
| `data/`                  | Data files used in analysis, when appropriate        |

# Key Takeaways

The main concepts to remember are:

1. **HTML** is the language browsers use to display web pages.
2. **Quarto** allows you to create web pages without writing large amounts of HTML manually.
3. A **`.qmd` file** combines writing, formatting, and executable code.
4. **YAML** controls document-level settings.
5. **Markdown** controls the structure and formatting of your written content.
6. **Code chunks** allow Quarto to execute R code during rendering.
7. **Render** builds the finished HTML output.
8. **Preview** lets you view the website locally on your own computer.
9. **Commit** saves a snapshot of your work in Git.
10. **Push** sends your commits to GitHub.
11. **GitHub Pages** publishes your website online.


Set up the portfolio workflow: made a repository from a template, cloned it with
GitHub Desktop, edited files in RStudio, rendered with Quarto, and published to
GitHub Pages.

## Key Concepts

Three places a file can live:

1. **My computer** — where I edit
2. **GitHub.com** — where it's stored
3. **My live site** — what the world sees

The loop: **Edit → Save → Render → Commit → Push → wait → refresh**

## Proving R Works

The chunk below runs when this page renders. If you can see a result underneath
it on your published site, R and Quarto are talking to each other correctly.

```{r}
sessionInfo()$R.version$version.string
```

## Reflections

What I want out of this course: I want to develop a stronger understanding of how Quarto, R, Git, GitHub, and GitHub Pages work together to create and publish data-driven websites. I also want to become more confident in using R for data analysis, visualization, and communicating results in a clear and professional way. By the end of the course, I hope to have a portfolio that demonstrates both my technical skills and my ability to communicate analytical findings. Most importantly, I want to develop skills that I can apply to future academic and professional projects.

One thing I'm nervous about:One thing I am nervous about is troubleshooting technical problems when my code, Quarto documents, or GitHub workflow does not work as expected. I am still becoming familiar with the different tools involved, so errors can sometimes be difficult to understand

## Questions I Still Have

How can I make my Quarto website look more professional and visually appealing?
What are the best practices for organizing data, code, images, and other files in a Quarto project?
How can I efficiently troubleshoot errors when rendering a .qmd file?
How can I use Git and GitHub effectively to track changes and manage different versions of my work?
What are some ways to make data visualizations more effective for communicating findings to a non-technical audience?
