---
title: "Assignment 3: Measuring Poverty in Philadelphia"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Emmanuel Sampson"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  freeze: false
  warning: true
  message: false
---

## Data and definition of poverty

How did poverty change across Philadelphia, and how much confidence can we
place in those changes? This analysis compares estimates from the American
Community Survey (ACS) for 2006–2010 and 2013–2017. It examines census tracts,
small statistical areas that help describe conditions within the city.
Each estimate represents a five-year period.

The poverty rate measures the percentage of people living below the poverty
level among those whose poverty status is determined by the survey. For
example, a rate of 25% means one in four people in that group lives below
the poverty level. The data come from the Census Bureau's
[poverty-status table](https://api.census.gov/data/2017/acs/acs5/groups/B17001.html).

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(tidyverse)
  library(tidycensus)
  library(sf)
})

poverty_variables <- c(total_pop = "B17001_001", poverty = "B17001_002")
fetch_poverty <- function(year) {
  suppressMessages(get_acs(
    geography = "tract", variables = poverty_variables,
    state = "PA", county = "Philadelphia", year = year,
    survey = "acs5", output = "wide", geometry = TRUE, moe_level = 90
  ))
}
```

```{r}
#| label: download-data
#| include: false
#| message: false
#| warning: false
#| results: hide
poverty_2010 <- fetch_poverty(2010)
poverty_2017 <- fetch_poverty(2017)
```

## Poverty rates and margins of error

The ACS surveys a sample of residents, so its estimates carry sampling
uncertainty. A margin of error (MOE) describes that uncertainty: a larger
MOE means a less precise estimate. A poverty rate of 25% with an MOE of
5 **percentage points** gives an approximate 90% confidence interval of
20%–30%. Across repeated samples, intervals calculated this way would
contain the true value about 90% of the time.

The calculations account for uncertainty in both the number of people below
the poverty level and the population used to calculate the rate, using the
[proportion method documented by tidycensus](https://walker-data.com/tidycensus/articles/margins-of-error.html).

```{r}
#| label: calculate-rates
calculate_rates <- function(data, period_label) {
  data %>%
    mutate(
      # Treat unavailable values and Census missing-data codes as missing.
      across(c(total_popE, total_popM, povertyE, povertyM),
             ~ replace(.x, !is.finite(.x) | .x < 0, NA_real_)),
      denominator = if_else(total_popE > 0 & !is.na(povertyE) &
                              povertyE <= total_popE, total_popE, NA_real_),
      poverty_rate = 100 * povertyE / denominator,
      poverty_rate_moe = 100 * moe_prop(
        povertyE, denominator, povertyM, total_popM
      ),
      lower_90 = pmax(0, poverty_rate - poverty_rate_moe),
      upper_90 = pmin(100, poverty_rate + poverty_rate_moe),
      period = period_label
    ) %>% select(-denominator)
}
poverty_2010 <- calculate_rates(poverty_2010, "2006-2010")
poverty_2017 <- calculate_rates(poverty_2017, "2013-2017")

poverty_2017 %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, poverty_rate_moe, lower_90, upper_90) %>%
  head(10) %>%
  knitr::kable(digits = 2,
               col.names = c("Tract ID", "Poverty rate (%)", "Margin of error (points)",
                             "90% lower bound (%)", "90% upper bound (%)"),
               caption = "Poverty estimates for ten tracts, 2013–2017")
```

## Reliability of tract estimates

To compare precision across tracts, this analysis expresses each margin of
error as a percentage of its estimated poverty rate. This **relative MOE**
is smaller when the estimate is more precise. For example, an MOE of
5 percentage points around a rate of 25% gives a relative MOE of 20%.

This analysis labels relative MOEs below 12% as reliable, 12% to less than
40% as somewhat reliable, and 40% or above as unreliable. These thresholds
are specific to this analysis. A relative margin of error cannot be
calculated when the estimated poverty rate is zero, because that would
require dividing by zero. These cases and missing data are listed separately.

```{r}
#| label: reliability
classify_reliability <- function(data) {
  data %>% mutate(
    moe_pct = 100 * poverty_rate_moe /
      if_else(poverty_rate > 0, poverty_rate, NA_real_),
    reliability = case_when(
      is.na(poverty_rate) | is.na(poverty_rate_moe) ~ "No valid estimate/MOE",
      poverty_rate == 0 ~ "Zero estimate: relative MOE undefined",
      moe_pct < 12 ~ "Reliable",
      moe_pct < 40 ~ "Somewhat reliable",
      TRUE ~ "Unreliable"
    )
  )
}
poverty_2010 <- classify_reliability(poverty_2010) %>% st_transform(4326)
poverty_2017 <- classify_reliability(poverty_2017) %>% st_transform(4326)
poverty_change <- bind_rows(poverty_2010, poverty_2017)

reliability_summary <- poverty_change %>% st_drop_geometry() %>%
  count(period, reliability, name = "tracts") %>%
  group_by(period) %>% mutate(percent = 100 * tracts / sum(tracts)) %>%
  ungroup()
knitr::kable(reliability_summary, digits = 1,
             col.names = c("Period", "Reliability", "Number of tracts", "Share of tracts (%)"),
             caption = "Reliability by period (including unavailable estimates)")
```

Estimates labeled unreliable have large margins of error compared with the
estimated rates. This makes it difficult to rank those tracts confidently
by poverty rate. Even a modest margin of error can be large relative to a
very low poverty rate.

## Change between periods

Tracts are matched across periods using their Census geographic identifiers
(`GEOID`). The table below shows how many appear in both periods or only
one. Comparisons assume the matched tracts represent comparable areas;
matching identifiers alone does not verify that boundaries are identical.
The Census Bureau documents the geographic definitions in its
[ACS geography guidance](https://www.census.gov/programs-surveys/acs/geography-acs/geography-boundaries-by-year/2010.html).

```{r}
#| label: match-tracts
tract_data <- poverty_change %>% st_drop_geometry()
stopifnot(!anyDuplicated(tract_data[c("GEOID", "period")]))
coverage <- full_join(
  st_drop_geometry(poverty_2010) %>% transmute(GEOID, in_2010 = TRUE),
  st_drop_geometry(poverty_2017) %>% transmute(GEOID, in_2017 = TRUE),
  by = "GEOID"
) %>% mutate(coverage = case_when(
  is.na(in_2010) ~ "2017 only",
  is.na(in_2017) ~ "2010 only",
  TRUE ~ "Both periods"
))
knitr::kable(count(coverage, coverage, name = "tracts"),
             col.names = c("Data availability", "Number of tracts"))

poverty_wide <- tract_data %>%
  select(GEOID, period, poverty_rate, poverty_rate_moe) %>%
  pivot_wider(id_cols = GEOID, names_from = period,
              values_from = c(poverty_rate, poverty_rate_moe))
```

The change is the later poverty rate minus the earlier rate. For example,
an increase from 20% to 25% is a rise of 5 percentage points. Assuming independent estimates for these
non-overlapping periods, the change MOE is the square root of the sum of
the squared MOEs. A change is significant at approximately the 90% level
when its absolute value exceeds this MOE. Each tract is tested separately,
without adjusting for the number of tests. With many tests, some changes
may appear significant by chance. Failure to detect a change does not prove
poverty stayed constant.

```{r}
#| label: test-change
poverty_wide <- poverty_wide %>% mutate(
  change_pp = `poverty_rate_2013-2017` - `poverty_rate_2006-2010`,
  change_moe = sqrt(`poverty_rate_moe_2013-2017`^2 +
                      `poverty_rate_moe_2006-2010`^2),
  change_lower_90 = change_pp - change_moe,
  change_upper_90 = change_pp + change_moe,
  change_result = case_when(
    is.na(change_pp) | is.na(change_moe) ~ "Not estimable",
    change_lower_90 > 0 ~ "Significant increase",
    change_upper_90 < 0 ~ "Significant decrease",
    TRUE ~ "No significant change"
  )
)
change_summary <- count(poverty_wide, change_result, name = "tracts")
knitr::kable(change_summary,
             col.names = c("Change assessment", "Number of tracts"),
             caption = "Tract changes at the 90% confidence level")

poverty_wide %>% filter(!is.na(change_moe), !is.na(change_pp)) %>%
  arrange(desc(abs(change_pp))) %>%
  select(GEOID, change_pp, change_moe, change_result) %>% head(10) %>%
  knitr::kable(digits = 2,
               col.names = c("Tract ID", "Change (points)", "Margin of error (points)",
                             "Change assessment"),
               caption = "Ten largest absolute estimated changes (percentage points)")
```

### Plot of change and its uncertainty

In the formulas below, $p$ is the poverty rate, $\Delta$ is the change
between periods, and $MOE$ is the margin of error. Rates and changes are
expressed in percentage points for these calculations.

$$
\Delta = p_{2013-2017} - p_{2006-2010}, \qquad
MOE_{\Delta} = \sqrt{MOE_{2013-2017}^{2} + MOE_{2006-2010}^{2}}.
$$

The graph shows every tract for which a change can be calculated, ordered from the
largest decrease to the largest increase. Each point is an estimated change;
each vertical line spans $\Delta \pm MOE_{\Delta}$. Both are in percentage
points. A confidence interval entirely above or below the dashed zero line
indicates a statistically significant change at 90% confidence. Intervals
that touch or cross zero do not establish a change at this confidence level.
The intervals use a normal approximation and are shown in full. For a few
tracts, large margins of error extend beyond the possible range of -100 to
100 percentage points. These wide intervals show how uncertain the estimates
are; changes beyond that range are not physically possible.

```{r}
#| label: fig-poverty-change
#| fig-width: 10
#| fig-height: 6
#| fig-dpi: 180
#| fig-cap: "Philadelphia tract poverty-rate changes, 2006–2010 to 2013–2017. Points show estimated changes; vertical lines show 90% confidence intervals for those changes. Tracts are ordered by estimated change, not geographic location."
change_plot_data <- poverty_wide %>%
  filter(is.finite(change_pp), is.finite(change_moe)) %>%
  arrange(change_pp, GEOID) %>%
  mutate(tract_rank = row_number())

ggplot(change_plot_data,
       aes(x = tract_rank, y = change_pp, color = change_result)) +
  geom_linerange(aes(ymin = change_lower_90, ymax = change_upper_90),
                 linewidth = 0.35, alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 1.2) +
  scale_color_manual(
    values = c("Significant decrease" = "#0072B2",
               "No significant change" = "#777777",
               "Significant increase" = "#D55E00"),
    breaks = c("Significant decrease", "No significant change",
               "Significant increase"),
    drop = FALSE
  ) +
  labs(
    title = "Change in poverty rates across Philadelphia census tracts",
    subtitle = "2013-2017 minus 2006-2010; intervals show uncertainty in the change",
    x = "Tracts ordered by estimated change",
    y = "Change in poverty rate (percentage points)",
    color = NULL,
    caption = paste(nrow(change_plot_data), "tracts shown;",
                    nrow(poverty_wide) - nrow(change_plot_data),
                    "tracts omitted because change cannot be estimated.")
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.caption = element_text(hjust = 0))
```

**Results.** Of `r nrow(poverty_wide)` census tracts across the two periods,
`r sum(poverty_wide$change_result == "Significant increase")` show a significant
increase and `r sum(poverty_wide$change_result == "Significant decrease")` show
a significant decrease at the 90% level.
There are `r sum(poverty_wide$change_result == "No significant change")` tracts
with no statistically detectable change and
`r sum(poverty_wide$change_result == "Not estimable")` without sufficient data.
These counts describe how many areas experienced detectable changes.
Because tracts have different populations, the counts alone do not show
whether Philadelphia's overall poverty rate increased or decreased.


